---
title: "Health Informatio System project "
author: "Bettinzoli Nicola"
date: "2024-06-18"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Introdution

The purpose of this report is to analyse the current state and the evolution of the air quality in the region of Lombardy. The ultimate objective is to identify which areas are more polluted 
  

# Library used

```{r library, ECHO = FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(tidyr)
})
```

# Datasets used


## Dati sensori aria


<!-- import of the dataset of the readings  -->
```{r}
df <- read.csv("./datasets/sensor_readings.csv")
str(df)
```


## Dati stazioni qualità dell'aria


<!-- import the dataset of the station used and look at its structure -->
```{r}
df_stations <- read.csv("./datasets/station_dataset.csv")
str(df_stations)
```

# Cleaning and manipuleting the datasets

## Check for NA records

<!-- check how many records in the dataset are NA  -->
```{r}
table(df$Stato, useNA = "always")
```
<!-- double check on the "Valore" field -->
```{r}
length(which(df$Valore == -9999))
```

<!-- drop Na rows -->
```{r, ECHO = FALSE}
df <- drop_na(df)
```

## Date feature

<!-- split the "Data" column into date and time -->
```{r}
dates <- df$Data
```

```{r}
dates <- format(as.POSIXct(dates, format="%d/%m/%Y %H:%M:%S"))
```

```{r}
times <- as.integer(format(as.POSIXct(dates), format = "%H:%M:%S"), format = "%H")
```

```{r}
dates <- as.Date(dates)
```

<!-- insert the new features into the dataset -->
```{r}
df$Dates <- dates

df$Times <- times
```

<!-- remove the useless column column -->
```{r}
df <- subset(df, select = -c(Data,idOperatore,Stato))
```

<!-- change the name of the id column, this for compatibility reason with the other dataset -->
```{r}
df <- rename(df,"IdSensore" = "idSensore")
```

<!-- check if all the ids of the first dataset exist in the second -->
```{r}
ids <- df$IdSensore
for (id in ids){
  if (!(id %in% df_stations$IdSensore))
    print("sensor "+id+" not present in the dataset")
}
  
```

<!-- join the two dataset based on the "IdSensore" feature -->
```{r}
df <- left_join(df,df_stations, by = "IdSensore")
```

<!-- remove some column that are not relevant -->
```{r}
df <- subset(df, select = -c(NomeStazione,Storico))
```

# Analyses


# Conclusions



