---
title: "Health Informatio System project "
author: "Bettinzoli Nicola"
date: "2024-06-18"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Introdution

The purpose of this report is to analyse the current state and the evolution of the air quality in the region of Lombardy. During this study i will analize the various concentration of the pollutants in the air and try to identify pattern or trend of the territory.  
  

# Library used

```{r library, ECHO = FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(tidyr)
})
```

# Datasets used


## Dati sensori aria

The main dataset utilized is the one containing all the sensor readings, this dataset can be downloaded from <https://www.dati.lombardia.it/Ambiente/Dati-sensori-aria-dal-2018/g2hp-ar79/about_data>. Let's try to have a look

<!-- import of the dataset of the readings  -->
```{r}
df <- read.csv("./datasets/sensor_readings.csv")
head(df)
```

And the structure
```{r}
str(df)
```


In this dataset each row rappresent a sensor reading, each reading is described by: The id of the sensor, the date, the value, the state and the id of the operator.


## Dati stazioni qualità dell'aria


<!-- import the dataset of the station used and look at its structure -->
```{r}
df_stations <- read.csv("./datasets/station_dataset.csv")
str(df_stations)
```

# Cleaning and manipuleting the datasets

## Check for NA records

<!-- check how many records in the dataset are NA  -->
```{r}
table(df$Stato, useNA = "always")
```
<!-- double check on the "Valore" field -->
```{r}
length(which(df$Valore == -9999))
```

<!-- drop Na rows -->
```{r, ECHO = FALSE}
df <- drop_na(df)
```

## Date feature

<!-- split the "Data" column into date and time -->
```{r}
dates <- df$Data
```

```{r}
dates <- format(as.POSIXct(dates, format="%d/%m/%Y %H:%M:%S"))
```

```{r}
times <- as.integer(format(as.POSIXct(dates), format = "%H:%M:%S"), format = "%H")
```

```{r}
dates <- as.Date(dates)
```

<!-- insert the new features into the dataset -->
```{r}
df$Dates <- dates

df$Times <- times
```

<!-- remove the useless column column -->
```{r}
df <- subset(df, select = -c(Data,idOperatore,Stato))
```

<!-- change the name of the id column, this for compatibility reason with the other dataset -->
```{r}
df <- rename(df,"IdSensore" = "idSensore")
```

<!-- check if all the ids of the first dataset exist in the second -->
```{r}
ids <- df$IdSensore
for (id in ids){
  if (!(id %in% df_stations$IdSensore))
    print("sensor "+id+" not present in the dataset")
}
  
```

<!-- join the two dataset based on the "IdSensore" feature -->
```{r}
df <- left_join(df,df_stations, by = "IdSensore")
```

<!-- remove some column that are not relevant -->
```{r}
df <- subset(df, select = -c(NomeStazione,Storico))
```

# Analyses


# Conclusions



